On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-03-28 manuscripts/PNAS/chazthm_pnas.tex
more sml update

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{Sparse linear coding, dictionary learning, compressed sensing}

@@

%  We introduce the following terminology to handle $\eta > 0$. 
\begin{definition}\label{maindef}
Fix $Y = \{ \mathbf{y}_1, \ldots, \mathbf{y}_N\} \subset \mathbb{R}^n$. We say $Y$ has a \textbf{$k$-sparse linear representation in $m$ dimensions} if there exists a matrix $\mathbf{A}$ and $k$-sparse $\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$ such that $\mathbf{y}_i = \mathbf{A}\mathbf{x}_i$ for all $i$. %Equivalently, $\mathbf{Y} = \mathbf{A}\mathbf{X}$ for $\mathbf{Y}$ and $\mathbf{X}$ having the $\mathbf{y}_i$ and $\mathbf{x}_i$ as their columns, respectively. 
This representation is \textbf{stable} if for every $\delta_1, \delta_2 \geq 0$, there exists $\varepsilon = \varepsilon(\delta_1, \delta_2) \geq 0$ (with $\varepsilon > 0$ when  $\delta_1, \delta_2 > 0$) such that if a matrix $\mathbf{B}$ and $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N \in \mathbb{R}^m$ satisfy:
\begin{align*}
\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon, \ \ \text{for all $i$},
\end{align*}
%
then there is some permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$ such that for all $i, j$:
\begin{align}\label{def1}
\|\mathbf{A}_j - (\mathbf{BPD})_j\|_2 \leq \delta_1 \ \ \text{and} \ \ \|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_i\|_1 \leq \delta_2.
\end{align}
\end{definition}
% Kato mentions a distance between n-tuples that minimizes the maximum difference over all possible orderings of one of the n-tuples.
% Make just 'sparse representation in m-dimensions? (i.e. with k unspecified?)

To see how Def. \ref{maindef} directly relates to Prob. \ref{InverseProblem}, suppose that $Y$
% \mbox{$Y = \{\mathbf{A} \mathbf{x}_1, \ldots, \mathbf{A}\mathbf{x}_N\}$} 
has a stable $k$-sparse linear representation in $m$ dimensions and fix $\delta_1, \delta_2$ to be the desired recovery accuracy in \eqref{def1}. Consider now any dataset $Z$ generated as in \eqref{LinearModel} with $\eta \leq \frac{1}{2} \varepsilon(\delta_1, \delta_2)$. Then from the triangle inequality, it follows that any matrix $\mathbf{B}$ and $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N$ solving Prob.~\ref{InverseProblem} are necessarily close to the original dictionary $\mathbf{A}$ and codes $\mathbf{x}_i$.

@@

In the next section, we give precise statements of our main results, which include an explicit form for $\varepsilon(\delta_1, \delta_2)$. We then prove our main theorem (Thm.~\ref{DeterministicUniquenessTheorem}) in Sec.~\ref{DUT} after stating some additional definitions and lemmas required for the proof, including our main tool from combinatorial matrix analysis (Lem.~\ref{MainLemma}). All other proofs are relegated to the Supplemental. 
%Our approach is a refinement of the one found in \cite{Hillar15} to handle noise and to reduce the sufficient number of samples.
%for all $k < m$ from $N=k{m \choose k}^2$ to $N = m + m(k-1){m \choose k}$. 
Finally, several applications are considered in our discussion, Sec.~\ref{Discussion}.

@@

To prepare for the statements of our main results, we first describe a quantitative version of the spark condition.  % We first explain how the spark condition \eqref{SparkCondition} relates to t
The \emph{lower bound} \cite{Grcar10} of a matrix $\mathbf{A} \in \mathbb R^{n \times m}$ is the largest number $\alpha$ such that \mbox{$\|\mathbf{A}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2$} for all $\mathbf{x} \in \mathbb{R}^m$. By compactness of the unit sphere, every injective linear map has a nonzero lower bound; hence, if $\mathbf{A}$ satisfies \eqref{SparkCondition}, then each submatrix formed from $2k$ of its columns or less has a nonzero lower bound. We therefore define the following domain-restricted lower bound of $\mathbf{A}$:
\begin{align*}
L_k(\mathbf{A}) := \frac{1}{\sqrt{k}}\max \{ \alpha : \|\mathbf{A}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2 \text{ for all $k$-sparse } \mathbf{x}\}.
\end{align*} 
Clearly, $L_k(\mathbf{A}) \geq L_{k'}(\mathbf{A})$ whenever $k < k'$, and for any $\mathbf{A}$ satisfying the spark condition, we have $L_{k'}(\mathbf{A}) > 0$ for all $k' \leq 2k$. Note that subtracting $\sqrt{k} L_k(\mathbf{A})$ [*** TODO: check is this correct ***] from one yields what is known in the compressed sensing literature as the (asymmetric) lower restricted isometry constant \cite{Blanchard2011, Foucart2009}.

% an important contribution of our work is to
Next, we identify our combinatorial criteria on the support sets of generating codes that allow for stable sparse 
representations.  Let $[m]$ denote the set $\{1, \ldots, m\}$, $2^{[m]}$ its power set, and ${[m] \choose k}$ the set of subsets of $[m]$ of size $k$. Let $\text{Span}\{\mathbf{v}_1, \ldots, \mathbf{v}_\ell\}$ be the $\mathbb{R}$-linear span of vectors $\mathbf{v}_1, \ldots, \mathbf{v}_\ell$. A vector $\mathbf{x}$ is then said to be \emph{supported} on $S \subseteq [m]$ when $\mathbf{x} \in \text{Span} \{\mathbf{e}_j\}_{j\in S}$, where $\mathbf{e}_j$ are the standard $m$-dimensional basis vectors We denote by $\mathbf{M}_j$ the $j$th column of a matrix $\mathbf{M}$ and similarly by $\mathbf{M}_S$ the sub-matrix formed by the columns of $\mathbf{M}$ indexed by $S$. We interpret $\text{Span}\{\mathbf{M}_\emptyset\} = \{\textbf{0}\}$, where $\emptyset$ is the empty set. We say a hypergraph $E \subseteq 2^{[m]}$ is of \textit{rank} $k$ when $|S| \leq k$ for every $S \in E$ (and \textit{$k$-uniform} when equality holds). We say $E$ is \emph{regular} when every element of $[m]$ is contained in exactly $\ell$ elements of $E$ for some $\ell > 0$ (for given $\ell$, we say $E$ is $\ell$-regular).

@@

\begin{theorem}\label{DeterministicUniquenessTheorem}
% Fix positive integers $k, \ell, m, m', n, p$ satisfying $k < m$ and $p \leq m \leq m' \leq (\ell m - p) / (\ell - 1)$. 
Fix integers $n, \ell$ and $k < m \leq \hat m$. Suppose $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ satisfies \eqref{SparkCondition} and that $k$-sparse \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} contain at least \mbox{$(k-1){\hat m \choose k}+1$} vectors supported in general linear position on each set in some $\ell$-regular $E \subseteq {[m] \choose k}$ satisfying the SIP.  There exists a constant $C_1 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_1$.

Every matrix $\mathbf{B} \in \mathbb{R}^{n \times \hat m}$ and $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N \in \mathbb{R}^{\hat m}$ with  \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon$} for all $i$ necessarily satisfies:
\begin{align}\label{Cstable}
\|\mathbf{A}_{\phi(j)} - (\mathbf{B}_{J}\mathbf{PD})_j\|_2 \leq C_1 \varepsilon, \ \ \text{for $j = 1, \ldots, |J|$},
\end{align}
%
for some $J \subseteq [\hat m]$ of size \mbox{$\lfloor \hat m - \ell(\hat m - m) \rfloor$}, injective map $\phi$, permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$.
%\footnote{We note that the condition $\varepsilon < L_2(\mathbf{A})$ above is necessary. When \mbox{$\mathbf{A}$ = $I$} and $\mathbf{x}_i = \mathbf{e}_i$, it turns out that $C_1 = 1$ and there is a $\mathbf{B}$ and $1$-sparse $\mathbf{\hat x}_i$ with $\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon$ violating \eqref{Cstable}.}

Furthermore if $\varepsilon < L_{2k}(\mathbf{A}) / C_1$ then $\mathbf{B}_J$ also satisfies \eqref{SparkCondition} with $L_{2k}(\mathbf{B}_J\mathbf{PD}) \geq L_{2k}(\mathbf{A}) - C_1 \varepsilon$ and we in fact have [ ** TODO ** ]:
\begin{align}\label{b-PDa}
\|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_i\|_1 &\leq  \left( \frac{ 1+C_1 \|\mathbf{x}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right) \varepsilon,
 \ \ \text{for all $i \in [N]$}.
\end{align}
\end{theorem}

In other words, the smaller the difference $\hat m - m$, the more columns and coefficients of the original $n \times m$ dictionary $\mathbf{A}$ and $m$-dimensional codes $\mathbf{x}_i$ contained (up to noise) in the appropriately scaled $n \times \hat m$ dictionary $\mathbf{B}$ and $\hat m$-dimensional codes $\mathbf{\hat x}_i$. In particular, if $\hat m = m$ then all columns of $\mathbf{A}$ and coefficients of each $\mathbf{x}_i$ recoverable, as per Def.~\ref{maindef}. For clarity of exposition, we delay defining the (explicit) constant $C_1 = C_1(\mathbf{A}, \{\mathbf{x}_i\}_{i=1}^N, E)$ until Section \ref{DUT} (\eqref{Cdef}), since it requires additional definitions not critical to the statement of our results.

@@

Since $f$ is a real analytic function, it is enough for one substitution of real numbers to satisfy $f(\mathbf{A}) \neq 0$ to conclude that its zeroes form a set with measure zero. Hence, almost every $n \times m$ real matrix $\mathbf{A}$ satisfies \eqref{SparkCondition} provided \eqref{CScondition} holds for a value of $\gamma$ for some distribution. We remark that the precise relationship between $m$, $n$, and $k$ guaranteeing that $f$ is not identically zero is a challenging problem in real algebraic geometry. In any case, we set $\gamma_0$ to be the smallest known such $\gamma$.

@@

Our first step is to demonstrate how dictionary recovery \eqref{Cstable} already implies sparse code recovery \eqref{b-PDa} when $\varepsilon < L_{2k}(\mathbf{A}) / C_1$. For all $2k$-sparse $\mathbf{x} \in \mathbb{R}^m$, the triangle inequality gives \mbox{$\|(\mathbf{A}-\mathbf{BPD})\mathbf{x}\|_2  \leq C_1\varepsilon \|\mathbf{x}\|_1 \leq C_1 \varepsilon \sqrt{2k}  \|\mathbf{x}\|_2$}. Thus, 
\begin{align*}
\|\mathbf{BPD}\mathbf{x}\|_2 
&\geq | \|\mathbf{A}\mathbf{x}\|_2 - \|(\mathbf{A}-\mathbf{BPD})\mathbf{x}\|_2 | \\
&\geq \sqrt{2k} (L_{2k}(\mathbf{A}) -  C_1\varepsilon) \|\mathbf{x}\|_2,
\end{align*}
%
where we drop the absolute value due to the upper bound on $\varepsilon$. Hence, $L_{2k}(\mathbf{BPD}) \geq L_{2k}(\mathbf{A}) - C_1\varepsilon  > 0$ and \eqref{b-PDa} then follows from:
\begin{align*}
\|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_i \|_1
%&\leq \sqrt{2k} |\mathbf{x}_i - D^{-1}P^{\top}\mathbf{\hat x}_i\|_2 \\
&\leq \frac{\|\mathbf{BPD}(\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_i)\|_2}{L_{2k}(\mathbf{BPD})} \\
&\leq \frac{\|\mathbf{B}\mathbf{\hat x}_i - \mathbf{A}\mathbf{x}_i\|_2 + \|(\mathbf{A} - \mathbf{BPD})\mathbf{x}_i\|_2}{L_{2k}(\mathbf{BPD})} \\
&\leq \frac{\varepsilon (1+C_1 \|\mathbf{x}_i\|_1)}{L_{2k}(\mathbf{BPD})}.
%&\leq \frac{\varepsilon}{\sqrt{2k}} \left( \frac{ 1+C_1 \|\mathbf{x}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right).
%\leq \frac{\varepsilon }{\varepsilon_0 - \varepsilon} \left( C_1^{-1}+|\mathbf{x}_i\|_1 \right).
\end{align*}

The heart of the matter is therefore \eqref{Cstable}, which we now establish using the following approach. Pigeonholing the data into its set of possible supports with respect to the matrix $\mathbf{B}$, one sees that the column space of every $\mathbf{A}_S$ with $S \in E$ must be nearby (in a sense yet to be defined; see Def.~\ref{dDef}) the column space of some $n \times k$ submatrix of $\mathbf{B}$. \eqref{Cstable} is then immediately implied by our result in combinatorial matrix analysis (Lem.~\ref{MainLemma}).

@@

\begin{definition}\label{dDef}
For any point $\mathbf{u}$, let $\text{dist}(\mathbf{u}, V) := \inf \{\| \mathbf{u}-\mathbf{v} \|_2: \mathbf{v} \in V\}$ be the distance of $\mathbf{u}$ from the set $V$. For subspaces $U,V \subseteq \mathbb{R}^m$, we define: %[Kato p.197]
\begin{align}\label{d}
d(U,V) := \sup_{\mathbf{u} \in U, \|\mathbf{u}\|_2 = 1} \text{dist}(\mathbf{u},V).
\end{align}
Note that $d(U,\{\textbf{0}\}) = 1$ for $U \neq \{\mathbf{0}\}$, whereas if $U = \{\textbf{0}\}$ then $d$ has no meaning; in this case, set $d(\{\textbf{0}\},V) = 0$ for any $V$.
\end{definition}
%Note that the supremum is always achievable by a point in $U$ when $U, V \subseteq \mathbb{R}^m$.

@@

\begin{lemma}\label{MainLemma}
Fix integers $n$ and $k < m$, and suppose $\mathbf{A} \in \mathbb{R}^{n \times m}$ satisfies the spark condition \eqref{SparkCondition}. There exists a constant $C_2 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_2$. If for some  $\mathbf{B} \in \mathbb{R}^{n \times m}$ and regular $E \subseteq2^{[m]}$ satisfying the singleton intersection property there exists a size-preserving map $\pi: E \mapsto 2^{[m]}$ satisfying:
\begin{align}\label{GapUpperBound}
d(\text{Span}\{\mathbf{A}_{S}\}, \text{Span}\{\mathbf{B}_{\pi(S)}\}) \leq \varepsilon, \ \ \text{for all $S \in E$},
\end{align}
%
then there exists some permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$ such that:
\begin{align}\label{MainLemmaBPD}
\|\mathbf{A}_j - (\mathbf{BPD})_j\|_2 \leq C_2 \varepsilon, \indent \text{for } j \in [m].
\end{align}
\end{lemma}

@@

The constant $C_2$ in Lem.~\ref{MainLemma} is then given by:  % \footnote{Note that $C_2 > 0$ is well-defined since $r > 0$ by definition.}
\begin{align}
C_2(\mathbf{A}, E) := \frac{ 2^{|E|} \max_{i \in [m]} \|\mathbf{A}_i\|_2}{ \min_{F \subseteq E} r( \{ \text{Span}\{\mathbf{A}_{S}\} \}_{S \in F}) }.
\end{align}

The reader may verify that this definition leads to a value for $C_1$ consistent with that used in our proof of the case $k=1$ above.

\begin{proof}[Proof ($k < m$)] 
%We begin by showing that for every $S \in E$ there is some $S' \in {[m] \choose k}$ for which the distance $d(\text{Span}\{\mathbf{A}_S\}, \text{Span}\{\mathbf{B}_S'\})$ is controlled by $\varepsilon$. Fix $S \in E$. Since there are $(k-1){m \choose k}+1$ vectors $\mathbf{x}_i$ supported on $S$, the pigeon-hole principle implies that there exists some $S' \in {[m] \choose k}$ and some set of $k$ indices $J(S)$ such that the supports of all $\mathbf{x}_i$ and $\mathbf{\hat x}_i$ with $i \in J(S)$ are contained in $S$ and $S'$, respectively.

>>>>

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{Sparse linear coding, dictionary learning, uniqueness, compressed sensing}

@@

%  We introduce the following terminology to handle $\eta > 0$. 
\begin{definition}\label{maindef}
Fix $Y = \{ \mathbf{y}_1, \ldots, \mathbf{y}_N\} \subset \mathbb{R}^n$. We say $Y$ has a \textbf{$k$-sparse linear representation in $m$ dimensions} if there exists a matrix $\mathbf{A}$ and $k$-sparse $\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$ such that $\mathbf{y}_i = \mathbf{A}\mathbf{x}_i$ for all $i$. %Equivalently, $\mathbf{Y} = \mathbf{A}\mathbf{X}$ for $\mathbf{Y}$ and $\mathbf{X}$ having the $\mathbf{y}_i$ and $\mathbf{x}_i$ as their columns, respectively. 
This representation is \textbf{stable} if for every $\delta_1, \delta_2 \geq 0$, there exists $\varepsilon(\delta_1, \delta_2) \geq 0$ (with $\varepsilon > 0$ when  $\delta_1, \delta_2 > 0$) such that if $\mathbf{B}$ and $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N \in \mathbb{R}^m$ satisfy:
\begin{align*}
\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon, \ \ \text{for all $i$},
\end{align*}
%
then there is some permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$ such that for all $i, j$:
\begin{align}\label{def1}
\|\mathbf{A}_j - (\mathbf{BPD})_j\|_2 \leq \delta_1 \ \ \text{and} \ \ \|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_i\|_1 \leq \delta_2.
\end{align}
\end{definition}
% Kato mentions a distance between n-tuples that minimizes the maximum difference over all possible orderings of one of the n-tuples.
% Make just 'sparse representation in m-dimensions? (i.e. with k unspecified?)

To see how Def. \ref{maindef} directly relates to Prob. \ref{InverseProblem}, suppose that $Y$
% \mbox{$Y = \{\mathbf{A} \mathbf{x}_1, \ldots, \mathbf{A}\mathbf{x}_N\}$} 
has a stable $k$-sparse linear representation and fix $\delta_1, \delta_2$ to be the desired recovery accuracy in \eqref{def1}. Consider now any dataset $Z$ generated as in \eqref{LinearModel} with $\eta \leq \frac{1}{2} \varepsilon(\delta_1, \delta_2)$. Then from the triangle inequality, it follows that any matrix $\mathbf{B}$ and $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N$ solving Prob.~\ref{InverseProblem} are necessarily close to the original dictionary $\mathbf{A}$ and codes $\mathbf{x}_i$.

@@

In the next section, we give precise statements of our main results, which include an explicit form for $\varepsilon(\delta_1, \delta_2)$. We then prove our main theorem (Thm.~\ref{DeterministicUniquenessTheorem}) in Sec.~\ref{DUT} after stating some additional definitions and lemmas required for the proof, including our main tool from combinatorial matrix analysis (Lem.~\ref{MainLemma}). All other proofs are relegated to the Supplemental. 
%Our approach is a refinement of the one found in \cite{Hillar15} to handle noise and to reduce the sufficient number of samples.
%for all $k < m$ from $N=k{m \choose k}^2$ to $N = m + m(k-1){m \choose k}$. 
Finally, we present several applications in Discussion Sec.~\ref{Discussion}.

@@

To prepare for the statements of our main results, we first describe a quantitative version of the spark condition.  % We first explain how the spark condition \eqref{SparkCondition} relates to t
The \emph{lower bound} \cite{Grcar10} of a matrix $\mathbf{A} \in \mathbb R^{n \times m}$ is the largest number $\alpha$ such that \mbox{$\|\mathbf{A}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2$} for all $\mathbf{x} \in \mathbb{R}^m$. By compactness of the unit sphere, every injective linear map has a nonzero lower bound; hence, if $\mathbf{A}$ satisfies \eqref{SparkCondition}, then each submatrix formed from $2k$ of its columns or less has a nonzero lower bound. We therefore define the following domain-restricted lower bound of $\mathbf{A}$:
\begin{align*}
L_k(\mathbf{A}) := \frac{1}{\sqrt{k}}\max \{ \alpha : \|\mathbf{A}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2 \text{ for all $k$-sparse } \mathbf{x}\}.
\end{align*} 
Clearly, $L_k(\mathbf{A}) \geq L_{k'}(\mathbf{A})$ whenever $k < k'$, and for any $\mathbf{A}$ satisfying the spark condition, we have $L_{k'}(\mathbf{A}) > 0$ for all $k' \leq 2k$. Note that the quantity $1 - \sqrt{k} L_k(\mathbf{A})$ [*** TODO: check is this correct ***] is known in the CS literature as the (asymmetric) lower restricted isometry constant \cite{Blanchard2011, Foucart2009}.

% an important contribution of our work is to
Next, we identify the combinatorial criteria on support sets of generating codes allowing for stable sparse 
representations.  Let $[m]$ denote the set $\{1, \ldots, m\}$, its power set by $2^{[m]}$, and ${[m] \choose k}$ the set of subsets of $[m]$ of size $k$. Let $\text{\rm Span}\{\mathbf{v}_1, \ldots, \mathbf{v}_\ell\}$ be the $\mathbb{R}$-linear span of vectors $\mathbf{v}_1, \ldots, \mathbf{v}_\ell$. A vector $\mathbf{x}$ is said to be \emph{supported} on $S \subseteq [m]$, written supp$(\mathbf{x}) = S$, when $\mathbf{x} \in \text{\rm Span} \{\mathbf{e}_j\}_{j\in S}$, where $\mathbf{e}_j$ are the standard basis elements in $\mathbb R^m$.  We denote by $\mathbf{M}_j$ the $j$th column of a matrix $\mathbf{M}$ and similarly by $\mathbf{M}_S$ the sub-matrix formed by the columns of $\mathbf{M}$ indexed by $S$ ($\text{\rm Span}\{\mathbf{M}_\emptyset\} := \{\textbf{0}\}$).  We say a hypergraph $E \subseteq 2^{[m]}$ is of \textit{rank} $k$ when every $S \in E$ has cardinality $|S| \leq k$ (and \textit{$k$-uniform} when equality holds). We say $E$ is \emph{regular} when every element of $[m]$ is contained in exactly $\ell$ elements of $E$ for some $\ell > 0$ (for given $\ell$, we say $E$ is $\ell$-regular).

@@

\begin{theorem}\label{DeterministicUniquenessTheorem}
% Fix positive integers $k, \ell, m, m', n, p$ satisfying $k < m$ and $p \leq m \leq m' \leq (\ell m - p) / (\ell - 1)$. 
Fix integers $n, \ell$ and $k < m \leq \hat m$. Suppose $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ satisfies the spark condition \eqref{SparkCondition} and that $k$-sparse \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} contain at least \mbox{$(k-1){\hat m \choose k}+1$} vectors supported in general linear position on each set in some $\ell$-regular $E \subseteq {[m] \choose k}$ satisfying the SIP.  There exists a constant $C_1 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_1$.

Every matrix $\mathbf{B} \in \mathbb{R}^{n \times \hat m}$ and $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N \in \mathbb{R}^{\hat m}$ with  \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon$} for all $i$ necessarily satisfies:
\begin{align}\label{Cstable}
\|\mathbf{A}_{\phi(j)} - (\mathbf{B}_{J}\mathbf{PD})_j\|_2 \leq C_1 \varepsilon, \ \ \text{for $j = 1, \ldots, |J|$},
\end{align}
%
for some $J \subseteq [\hat m]$ of size \mbox{$\lfloor \hat m - \ell(\hat m - m) \rfloor$}, injective map $\phi$, permutation matrix $\mathbf{P}$, and invertible diagonal matrix $\mathbf{D}$.
%\footnote{We note that the condition $\varepsilon < L_2(\mathbf{A})$ above is necessary. When \mbox{$\mathbf{A}$ = $I$} and $\mathbf{x}_i = \mathbf{e}_i$, it turns out that $C_1 = 1$ and there is a $\mathbf{B}$ and $1$-sparse $\mathbf{\hat x}_i$ with $\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon$ violating \eqref{Cstable}.}

Furthermore, if $\varepsilon < L_{2k}(\mathbf{A}) / C_1$, then $\mathbf{B}_J$ also satisfies \eqref{SparkCondition} with $L_{2k}(\mathbf{B}_J\mathbf{PD}) \geq L_{2k}(\mathbf{A}) - C_1 \varepsilon$, and:
\begin{align}\label{b-PDa}
\|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_i\|_1 &\leq  \left( \frac{ 1+C_1 \|\mathbf{x}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right) \varepsilon,
 \ \ \text{for all $i \in [N]$}.
\end{align}
\end{theorem}

In other words, the smaller the difference $\hat m - m$, the more columns and coefficients of the original $n \times m$ dictionary $\mathbf{A}$ and $m$-dimensional codes $\mathbf{x}_i$ contained (up to noise) in the appropriately scaled $n \times \hat m$ dictionary $\mathbf{B}$ and $\hat m$-dimensional codes $\mathbf{\hat x}_i$. In particular, when $\hat m = m$ all columns of $\mathbf{A}$ and coefficients of each $\mathbf{x}_i$ recoverable, as per Def.~\ref{maindef}. For expositional clarity, we delay defining the (explicit) constant $C_1 = C_1(\mathbf{A}, \{\mathbf{x}_i\}_{i=1}^N, E)$ until Section \ref{DUT} (\eqref{Cdef}), since it requires additional definitions not critical to the statement of our results.

@@

Since $f$ is a real analytic function, one substitution of real numbers with $f(\mathbf{A}) \neq 0$ implies that its zeroes form a set with measure zero. Hence, almost every $n \times m$ real matrix $\mathbf{A}$ satisfies \eqref{SparkCondition} provided \eqref{CScondition} holds for a value of $\gamma$ for some distribution. We remark that the precise relationship between $m$, $n$, and $k$ guaranteeing that $f$ is not identically zero is a challenging problem in real algebraic geometry. In any case, we set $\gamma_0$ to be the smallest known such $\gamma$.

@@

Our first step is to show how dictionary recovery \eqref{Cstable} already implies sparse code recovery \eqref{b-PDa} when $\varepsilon < L_{2k}(\mathbf{A}) / C_1$. For all $2k$-sparse $\mathbf{x} \in \mathbb{R}^m$, the triangle inequality gives \mbox{$\|(\mathbf{A}-\mathbf{BPD})\mathbf{x}\|_2  \leq C_1\varepsilon \|\mathbf{x}\|_1 \leq C_1 \varepsilon \sqrt{2k}  \|\mathbf{x}\|_2$}. Thus, 
\begin{align*}
\|\mathbf{BPD}\mathbf{x}\|_2 
&\geq | \|\mathbf{A}\mathbf{x}\|_2 - \|(\mathbf{A}-\mathbf{BPD})\mathbf{x}\|_2 | \\
&\geq \sqrt{2k} (L_{2k}(\mathbf{A}) -  C_1\varepsilon) \|\mathbf{x}\|_2,
\end{align*}
%
where we drop the absolute value due to the upper bound on $\varepsilon$. Hence, $L_{2k}(\mathbf{BPD}) \geq L_{2k}(\mathbf{A}) - C_1\varepsilon  > 0$ and \eqref{b-PDa} then follows from:
\begin{align*}
\|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_i \|_1
%&\leq \sqrt{2k} |\mathbf{x}_i - D^{-1}P^{\top}\mathbf{\hat x}_i\|_2 \\
&\leq \frac{\|\mathbf{BPD}(\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_i)\|_2}{L_{2k}(\mathbf{BPD})} \\
&\leq \frac{\|\mathbf{B}\mathbf{\hat x}_i - \mathbf{A}\mathbf{x}_i\|_2 + \|(\mathbf{A} - \mathbf{BPD})\mathbf{x}_i\|_2}{L_{2k}(\mathbf{BPD})} \\
&\leq \frac{\varepsilon (1+C_1 \|\mathbf{x}_i\|_1)}{L_{2k}(\mathbf{BPD})}.
%&\leq \frac{\varepsilon}{\sqrt{2k}} \left( \frac{ 1+C_1 \|\mathbf{x}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right).
%\leq \frac{\varepsilon }{\varepsilon_0 - \varepsilon} \left( C_1^{-1}+|\mathbf{x}_i\|_1 \right).
\end{align*}

The heart of the matter is therefore \eqref{Cstable}, which we now establish using the following approach. Pigeonholing the data into its set of possible supports with respect to the matrix $\mathbf{B}$, one sees that the column space of every $\mathbf{A}_S$ with $S \in E$ must be nearby (in a sense yet to be defined; see Def.~\ref{dDef}) the column space of some $n \times k$ submatrix of $\mathbf{B}$. \eqref{Cstable} then follows from our result in combinatorial matrix analysis (Lem.~\ref{MainLemma}).

@@

\begin{definition}\label{dDef}
For any point $\mathbf{u}$, let $\text{\rm dist}(\mathbf{u}, V) := \inf \{\| \mathbf{u}-\mathbf{v} \|_2: \mathbf{v} \in V\}$ be the distance of $\mathbf{u}$ from the set $V$. For subspaces $U,V \subseteq \mathbb{R}^m$, we define: %[Kato p.197]
\begin{align}\label{d}
d(U,V) := \sup_{\mathbf{u} \in U, \|\mathbf{u}\|_2 = 1} \text{\rm dist}(\mathbf{u},V).
\end{align}
Note that $d(U,\{\textbf{0}\}) = 1$ for $U \neq \{\mathbf{0}\}$, whereas if $U = \{\textbf{0}\}$ then $d$ has no meaning; in this case, set $d(\{\textbf{0}\},V) = 0$ for any $V$.
\end{definition}
%Note that the supremum is always achievable by a point in $U$ when $U, V \subseteq \mathbb{R}^m$.

@@

\begin{lemma}\label{MainLemma}
Fix integers $n$ and $k < m$, and suppose $\mathbf{A} \in \mathbb{R}^{n \times m}$ satisfies the spark condition \eqref{SparkCondition}. There exists a constant $C_2 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_2$. If for some  $\mathbf{B} \in \mathbb{R}^{n \times m}$ and regular $E \subseteq2^{[m]}$ satisfying the singleton intersection property there exists a size-preserving map $\pi: E \mapsto 2^{[m]}$ satisfying:
\begin{align}\label{GapUpperBound}
d(\text{\rm Span}\{\mathbf{A}_{S}\}, \text{\rm Span}\{\mathbf{B}_{\pi(S)}\}) \leq \varepsilon, \ \ \text{for all $S \in E$},
\end{align}
%
then there exists some permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$ such that:
\begin{align}\label{MainLemmaBPD}
\|\mathbf{A}_j - (\mathbf{BPD})_j\|_2 \leq C_2 \varepsilon, \indent \text{for } j \in [m].
\end{align}
\end{lemma}

@@

The constant $C_2$ in Lem.~\ref{MainLemma} is then given by:  % \footnote{Note that $C_2 > 0$ is well-defined since $r > 0$ by definition.}
\begin{align}
C_2(\mathbf{A}, E) := \frac{ 2^{|E|} \max_{i \in [m]} \|\mathbf{A}_i\|_2}{ \min_{F \subseteq E} r( \{ \text{\rm Span}\{\mathbf{A}_{S}\} \}_{S \in F}) }.
\end{align}

Note that this definition leads to a value for $C_1$ consistent with that used in our proof of the case $k=1$ above.

\begin{proof}[Proof ($k < m$)] 
%We begin by showing that for every $S \in E$ there is some $S' \in {[m] \choose k}$ for which the distance $d(\text{\rm Span}\{\mathbf{A}_S\}, \text{\rm Span}\{\mathbf{B}_S'\})$ is controlled by $\varepsilon$. Fix $S \in E$. Since there are $(k-1){m \choose k}+1$ vectors $\mathbf{x}_i$ supported on $S$, the pigeon-hole principle implies that there exists some $S' \in {[m] \choose k}$ and some set of $k$ indices $J(S)$ such that the supports of all $\mathbf{x}_i$ and $\mathbf{\hat x}_i$ with $i \in J(S)$ are contained in $S$ and $S'$, respectively.
